Non-word repetition in Yélî Dnye

### 2020-09-17 manuscript.Rmd
added effect sizes and info on semantic errors

@@

- add information about proportion of mispronunciations that are real based on other papers
- add effect sizes to all analyses
- ? correct treatment of words whose mispronunciation is an English word:
    mispronunciation meaning
361              two     two
362              two     two
879          no-pinêt                                          English: no peanut
908            nodedi                                           English: no daddy
933          no--dadi                                           English: no daddy
949           petrumi                                             English: "true"
950           petrumi                                             English: "true"
951           petrumi                                             English: "true"
1373              wee     English: where
1374              wee     English: where
1375              wee     English: where
1379              wee     English: where
1380              wee     English: where
1388              wee     English: where
1394              wee     English: where
1396              wee     English: where
1398              wee     English: where
1399              wee     English: where
1402              wat      English: what
1405              wat      English: what
1411              wee     English: where

@@

@@

A first list of candidate items was generated in 2018 by selecting simple consonants ("p",	"t",	"d",	"k",	"m",	"n",	"w",	"y") and vowels ("i", "o","u", "a","e") that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences. These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary. The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words. Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.

@@

```{r Fig1-first_vs_others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="NWR scores for individual participants averaging separately their first attempts and all other attempts."}

@@

plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="NWR scores (%)")
for(id in levels(factor(pwcm_att$ID))) lines(myx[pwcm_att$ID==id],pwcm_att$Word[pwcm_att$ID==id],lty=2)
axis(1,at=1:2,labels=c("first","others"))

t.test(pwcm_att$Word~pwcm_att$Attempt,paired=T)->myt
```

We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions. We excluded `r excl` children who did not have data for one of these two types. As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions. Surprisingly, NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`. Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.

@@

The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%). Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%). The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted. Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.

@@

In this analysis, we were interested in variation in NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world. To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not. The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID & target ID as random variables.

```{r xling-freq-ana}
trnsc$avg_fr_s=scale(trnsc$avg_fr)
trnsc$age.rounded_s=scale(trnsc$age.rounded)
xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
mycoeff=round(summary(xl_fr)$coefficients,3)
#summary(xl_fr)
#plot(xl_fr)
```

We could include 885 observations (**why -- shouldn't it be 1k??**), from 41 children producing in any given trial one of 41 potential target words. The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std. Error"]`, p = `r mycoeff["avg_fr_s","Pr(>|z|)"]`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2. Age was also a significant predictor  (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std. Error"]`, p = `r mycoeff["age.rounded_s","Pr(>|z|)"]`).

@@

@@

We next checked whether this may actually be due to frequency of the sounds within the language. We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early], and applied the natural logarithm. As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses). Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, =`r round(freq_cor$p.value,3)`). Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age. As before, the model contained random slopes for both child ID and target. In this model, both cross-linguistic frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std. Error"]`, p = `r both_coeff["avg_fr_s","Pr(>|z|)"]`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std. Error"]`, p = `r both_coeff["age.rounded_s","Pr(>|z|)"]`) were significant predictors of word-level NWR scores, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std. Error"]`, p = `r both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"]`) was not.

<!-- a2m should we point out that the coefficient for cross-ling frequency is reduced from .77 to .67, so indirectly corpus frequency does affect a little bit? Or making things too complicated, and just say "so corpus frequency does not explain away these results" -->

```{r diffic,echo=F}
nasalVs=c(":a", ":â" ,":ê", ":i" ,":o")
oralVs=c("a", "â","e",  "é" ,"ê", "i","î" ,"o","ó","u")
simplexCs=c("m","n","l","w","y","v","d","g","p","t","k","f","h","ch")
complexCs=c("dp","tp","kp","km","kn","mb","gh","lv")

@@

@@

**use proportion of correctly produced rather than number; DONE**

**a2m use first attempts rather than the whole data set: having second thoughts. To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers & it is more consistent. BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros... CF Option 3 further down**

**a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?**

@@

Finally, we looked into the frequency with which mispronunciations resulted in real words. Two thirds of them were: `r round(sum(!is.na(trnsc$meaning) & trnsc$meaning != "0")/sum(!is.na(trnsc$mispronunciation))*100)`%.

@@

@@

Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole word scoring. This is the most commonly reported type of score, but it is also the least forgiving. The pattern is less marked when other two scores are used, which are less sensitive to errors.

```{r tablength, results = "asis"}
birthorder.tab=NULL
for(myvar in c("Word","Phoneme","NLD")){
  birthorder.tab=cbind(birthorder.tab,
              paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
              " (",round(aggregate(by_length[,myvar],by=list(by_length$Length),sd)$x),")"))
}
rownames(birthorder.tab)<-levels(by_length$Length)
colnames(birthorder.tab)<-c("Word","Phoneme","NLD")
apa_table(birthorder.tab,caption="NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
```

>>>>

cohensd=function(m,sd){round(m/sd,3)}
cohensd2=function(m1,sd1,m2,sd2){round((m1-m2)/sqrt((sd1^2+sd2^2)/2),3)}
```

@@

- ? correct treatment of words whose mispronunciation is an English word:
    mispronunciation meaning
361              two     two
362              two     two
879          no-pinêt                                          English: no peanut
908            nodedi                                           English: no daddy
933          no--dadi                                           English: no daddy
949           petrumi                                             English: "true"
950           petrumi                                             English: "true"
951           petrumi                                             English: "true"
1373              wee     English: where
1374              wee     English: where
1375              wee     English: where
1379              wee     English: where
1380              wee     English: where
1388              wee     English: where
1394              wee     English: where
1396              wee     English: where
1398              wee     English: where
1399              wee     English: where
1402              wat      English: what
1405              wat      English: what
1411              wee     English: where

@@

# a2m I get a warning:
#   Warning message:
# Vectorized input to `element_text()` is not officially supported.
# Results may be unexpected or may change in future versions of ggplot2.

@@

A first list of candidate items was generated in 2018 by selecting simple consonants ("p",	"t",	"d",	"k",	"m",	"ń",	"w",	"y") 
and vowels ("i", "o","u", "a","e") that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences. These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary. The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words. Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.

@@

```{r Fig1-first_vs_others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts."}

@@

plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="NWR scores (% words)")
for(id in levels(factor(pwcm_att$ID))) lines(myx[pwcm_att$ID==id],pwcm_att$Word[pwcm_att$ID==id],lty=2)
axis(1,at=1:2,labels=c("first","others"))

t.test(pwcm_att$Word~pwcm_att$Attempt,paired=T)->myt

We first checked whether whole-item NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions. We excluded `r excl` children who did not have data for one of these two types. As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions. Surprisingly, whole-item NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`. Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.

@@

The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`. Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%), Cohen's d = `r cohensd(mean(overall$Phoneme),sd(overall$Phoneme))`. The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted. Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.

@@

In this analysis, we were interested in variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world. To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not. The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID & target ID as random variables.

```{r xling-freq-ana}
trnsc$avg_fr_s=scale(trnsc$avg_fr)
trnsc$age.rounded_s=scale(trnsc$age.rounded)
xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
mycoeff=round(summary(xl_fr)$coefficients,3)
#summary(xl_fr)
#plot(xl_fr)

aggregate(trnsc$correct*100,by=list(trnsc$avg_fr_s),mean)->xl_fr_tab
colnames(xl_fr_tab)<-c("xlf","Word")
xlfr_r=cor.test(xl_fr_tab$xlf,xl_fr_tab$Word)
```

We could include `r summary(xl_fr)$devcomp$dims["N"]` observations, from `r summary(xl_fr)$ngrps["file"]` children producing in any given trial one of `r summary(xl_fr)$ngrps["target"]` potential target words. The analysis revealed a main effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std. Error"]`, p = `r mycoeff["age.rounded_s","Pr(>|z|)"]`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std. Error"]`, p = `r mycoeff["avg_fr_s","Pr(>|z|)"]`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2. Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r r(`r length_r$parameter`) = `r round(xlfr_r$estimate,3)`.

@@

aggregate(trnsc$correct*100,by=list(trnsc$avg_fr_cor_ty_s),mean)->cor_fr_tab
colnames(cor_fr_tab)<-c("corf","Word")

@@

We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence is actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat. We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^2] As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses). Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, =`r round(freq_cor$p.value,3)`). Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = `r round(cor.test(cor_fr_tab$corf,cor_fr_tab$Word)$estimate,3)`. Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age. As before, the model contained random slopes for both child ID and target. In this model, both cross-linguistic frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std. Error"]`, p = `r both_coeff["avg_fr_s","Pr(>|z|)"]`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std. Error"]`, p = `r both_coeff["age.rounded_s","Pr(>|z|)"]`) were significant predictors of whole-item  NWR scores, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std. Error"]`, p = `r both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"]`) was not.

[^2]: We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less convincing.

<!-- a2m should we point out that the coefficient for cross-ling frequency is reduced, so indirectly corpus frequency does explain a little bit of variance? Or making things too complicated, and just say "so corpus frequency does not explain away these results" -->

### Qualitative analyses of NWR mispronunciations

<!-- a2m I realized this section did not belong together with the xling freq section above - are you ok with this title? -->

```{r diffic,echo=F}
nasalVs=c(":a", ":â" ,":ê", ":i" ,":o")
oralVs=c("a", "â","e",  "é" ,"ê", "i","î" ,"o","ó","u")
simplexCs=c("m","ń","l","w","y","v","d","g","p","t","k","f","h","ch")
complexCs=c("dp","tp","kp","km","kn","mb","gh","lv")

@@

#this means: divide each column by corresponding number in denominator vector
#sub_pc has the same shape as the original sub_tab, namely targets as rows, realization as columns
sub_pc=round(sub_tab/c(denom_vec)*100,3)

del_pc=round(del_tab/denom_vec*100,3)
mat_pc=round(match_tab/denom_vec*100,3)

@@

<!-- a2m I think it's weird to put effect sizes here, right? technically I could do something like cohen's d based on mean and sd of the percentage of tokens... -->

<!-- **a2m regarding the use first attempts rather than the whole data set, I'm having second thoughts. To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers & it is more consistent. BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros... CF Option 3 further down** -->

<!-- **a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?** -->

@@

Finally, we looked into the frequency with which mispronunciations resulted in real words. Two thirds of them were: `r round(sum(!is.na(trnsc$meaning) & trnsc$meaning != "0")/sum(!is.na(trnsc$mispronunciation))*100)`%. This is a type of analysis that is seldom reported, and we could only find one comparison point: @castro1998 found that illiterate adults NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate adults did so in only `r round(2/117*100,2)`%. The percentage we observe here is much higher than this, but we do not know whether age, language, or even test structure explains this difference.

<!-- a2m I added info above - is it weird to talk about this here? -->

@@

aggregate(trnsc$correct*100,by=list(trnsc$type),mean)->by_length_no_child
colnames(by_length_no_child)<-c("Length","Word")
by_length_no_child=by_length_no_child[-1,]
length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)
```

@@

Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole word scoring. This is the most commonly reported type of score, but it is also the least forgiving. The pattern is less marked when other two scores are used, which are less sensitive to errors. Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(`r length_r$parameter`) = `r round(length_r$estimate,3)`.
<!-- a2m it doesn't make sense to do this, right, with only 3 items? -->

```{r tablength, results = "asis"}
length.tab=NULL
for(myvar in c("Word","Phoneme","NLD")){
  length.tab=cbind(length.tab,
              paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
              " (",round(aggregate(by_length[,myvar],by=list(by_length$Length),sd)$x),")"))
}
rownames(length.tab)<-levels(by_length$Length)
colnames(length.tab)<-c("Word","Phoneme","NLD")
apa_table(length.tab,caption="NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
```
